---
title: "Template UNM reveal.js"
format: 
  unm-revealjs:
    slide-number: c/t
author:
  - name: Your Name
    orcid: 0000-0000-0000-0000
    email: alias@email.com
    # affiliations: Your Institution
# date: last-modified 
---

## Example slide

### This is Heading 3

Here we have some text that may run over several lines of the slide frame,
depending on how long it is.

- This is a bullet point
- first item 
    - A sub item

### Lists
1. first item
2. second item
   1. sub item

<!-- Next, we'll brief review some theme-specific components. -->

<!-- - Note that _all_ of the standard Reveal.js
[features](https://quarto.org/docs/presentations/revealjs/)
can be used with this theme, even if we don't highlight them here. -->

# Section Slide without Text {.section-slide}


# Section Slide with Text {.section-slidet}

This is a `{.section slide}` class with texts.

- This is useful for breaking up the presentation into sections.


## Additional theme classes

### Special classes for emphasis

- `.alert` class for default emphasis, e.g. [important note]{.alert}.
- `.fg` class for custom colour, e.g. [important note]{.fg style="--col: #e64173"}.
- `.bg` class for custom background, e.g. [important note]{.bg style="--col: #e64173"}.

### Cross-references

- `.button` class provides a Beamer-like button, e.g.
[[Summary]{.button}](#sec-summary)


## Code blocks

This is a regular code block in a slide:

```{r}
#| echo: true
# This is an R code block
x <- rnorm(100)
y <- 2*x + rnorm(100)
summary(lm(y ~ x))
``` 


## Code blocks {.small-slide}

This is a code slide with the class `{.small-slide}`, which is useful for code-heavy slides.

```{r}
#| echo: true
N <- 12 # Number of experimental units, Step 1
# Define experimental units and treatments
experimental_units <- 1:N ## 10 experimental units
treatments <- c("A", "B", "C") ## 3 treatments, Step 2

# Repeat treatments to match the number of units. The code covers both balanced and unbalanced designs.
treatments <- rep(treatments, length.out = N)  ## Step 3

# For a balanced design, we can also use:
# treatments <- rep(treatments, each = N/length(treatments))

# Shuffle (randomize) treatments randomly to assign to units, Step 4
set.seed(123) ## For reproducibility
randomized_treatments <- sample(treatments)

# Create a data frame with randomized assignments
randomization_crd <- data.frame(Unit = experimental_units, Treatment = randomized_treatments)  ## Step 5
print(randomization_crd)
```


## A Long Slide

Here is an example of a slide with a long list:

1.  Understand the rationale and structure of the CRD.
2.  Carry out a randomization procedure and interpret why coding treatments matters.
3.  Formulate and apply the one-way ANOVA model (including assumptions).
4.  Explore estimability, least squares estimation, and the distribution of
    estimators.
5.  Use the ANOVA test to decide if treatments differ significantly.
6.  Calculate sample size and power for planning experiments.
7.  Demonstrate R-based procedures for randomization, modeling, and ANOVA.
8.  Interpret the results of a CRD analysis.
9.  Apply the CRD to real-world research scenarios.
10. Understand the limitations and assumptions of the CRD.
11. Additional item
12. Additional item
13. Additional item
14. Additional item

## A Long Slide {.small-slide}

Here is an example of a slide with a long list and a smaller attribute class `{.small-slide}`:

1.  Understand the rationale and structure of the CRD.
2.  Carry out a randomization procedure and interpret why coding treatments matters.
3.  Formulate and apply the one-way ANOVA model (including assumptions).
4.  Explore estimability, least squares estimation, and the distribution of
    estimators.
5.  Use the ANOVA test to decide if treatments differ significantly.
6.  Calculate sample size and power for planning experiments.
7.  Demonstrate R-based procedures for randomization, modeling, and ANOVA.
8.  Interpret the results of a CRD analysis.
9.  Apply the CRD to real-world research scenarios.
10. Understand the limitations and assumptions of the CRD.
11. Additional item
12. Additional item
13. Additional item
14. Additional item

## A Long Slide {.tiny-slide}

Here is an example of a slide with a long list and a tiny attribute class `{.tiny-slide}`:

1.  Understand the rationale and structure of the CRD.
2.  Carry out a randomization procedure and interpret why coding treatments matters.
3.  Formulate and apply the one-way ANOVA model (including assumptions).
4.  Explore estimability, least squares estimation, and the distribution of
    estimators.
5.  Use the ANOVA test to decide if treatments differ significantly.
6.  Calculate sample size and power for planning experiments.
7.  Demonstrate R-based procedures for randomization, modeling, and ANOVA.
8.  Interpret the results of a CRD analysis.
9.  Apply the CRD to real-world research scenarios.
10. Understand the limitations and assumptions of the CRD.
11. Additional item
12. Additional item
13. Additional item
14. Additional item    

## Callouts Examples

Here are some examples of callout boxes:

::: {.callout-note}
This is a note callout box.
:::

::: {.callout-tip}
This is a tip callout box.
:::

::: {.callout-important}
This is an important callout box.
:::